---
title: "Water Quality Parameters: Inventory and Gaps"
execute:
  warning: false
  message: false
---

```{r setup}
library(dataRetrieval)
library(dplyr)
library(ggplot2)

# Fetch parameter data for all five HUCs
nushagak_hucs <- c("19030301", "19030302", "19030303", "19030304", "19030305")

message("Querying WQP for parameter observations...")
wq_data <- data.frame()

for (huc in nushagak_hucs) {
  temp <- readWQPdata(huc = huc, sampleMedia = "Water")
  temp$HUC <- huc
  wq_data <- rbind(wq_data, temp)
}

# Remove duplicates and NA characteristic names
wq_data <- wq_data %>%
  filter(!is.na(CharacteristicName))
```

## Parameter Frequency

The following parameters have been measured in the Nushagak watershed:

```{r param_freq}
param_freq <- wq_data %>%
  group_by(CharacteristicName) %>%
  summarize(
    Observations = n(),
    Stations = n_distinct(MonitoringLocationIdentifier),
    .groups = "drop"
  ) %>%
  arrange(desc(Observations)) %>%
  rename("Parameter" = CharacteristicName, "Measurements" = Observations, "Unique Stations" = Stations)

# Display top 20 parameters
knitr::kable(
  head(param_freq, 20),
  caption = "Most Frequently Measured Water Quality Parameters (Top 20)"
)
```

## Key Parameter Categories

### Physical Properties

- **Temperature**: Essential for salmon habitat assessment; continuous monitoring data available from USGS gages
- **Conductivity & Specific Conductance**: Indicates mineralization and ionic composition
- **Turbidity**: Affects light penetration; high values can indicate erosion or glacial discharge

### Chemical Properties

- **pH**: Affects aquatic chemistry and organism physiology
- **Dissolved Oxygen**: Critical for salmonid survival; cold, well-oxygenated water preferred
- **Alkalinity & Hardness**: Buffer capacity against acid inputs

### Nutrients

- **Nitrogen (various forms)**: Nitrate, nitrite, ammonia, total nitrogen
- **Phosphorus (various forms)**: Orthophosphate, total phosphorus
- (Limited frequency in existing data; opportunity for expanded monitoring)

### Trace Elements & Metals

- **Copper**: Abnormal levels can be toxic to aquatic life; relevant to mining activity concerns
- **Zinc, Lead, Cadmium, Chromium**: Other metals detected; relevance context-dependent
- **Molybdenum**: Associated with geochemistry and mining scenarios

### Biological

- **Algal and macroinvertebrate sampling**: Limited in frequency; primarily USGS and academic efforts

------------------------------------------------------------------------

## Data Gaps and Monitoring Opportunities

Based on the inventory, the following represent priorities for expanded collaborative monitoring:

1.  **Continuous Temperature Data**: Few sites have high-frequency temperature loggers; salmon thermal habitat monitoring would benefit from denser deployment.

2.  **Nutrients (N, P)**: Few measurements in existing WQP data; baseline nutrient levels are poorly characterized across most of the watershed.

3.  **Spatial Coverage in Wood River and Nuyakuk Drainages**: Fewer historical stations in these sub-basins; new effort could establish baseline for comparison.

4.  **Biological Metrics**: Macroinvertebrate and periphyton sampling is limited. Fish presence/absence and juvenile salmon surveys would complement chemistry data.

5.  **Seasonal and High-Flow Sampling**: Most data are episodic; systematic seasonal sampling would reveal temperature, discharge, and contaminant dynamics.

------------------------------------------------------------------------

These gaps inform the monitoring design recommendations in Part III.
