---
title: "Metals and Contaminants"
execute:
  warning: false
  message: false
---

```{r setup}
library(dataRetrieval)
library(dplyr)

nushagak_hucs <- c("19030301", "19030302", "19030303", "19030304")

# Key metals and trace elements
metals <- c("Copper", "Zinc", "Lead", "Cadmium", "Chromium", "Molybdenum", 
            "Arsenic", "Iron", "Manganese")

message("Querying WQP for trace metals...")
wq_data <- readWQPdata(huc = nushagak_hucs, sampleMedia = "Water")

metals_data <- wq_data %>%
  filter(CharacteristicName %in% metals) %>%
  filter(!is.na(ResultMeasureValue)) %>%
  select(CharacteristicName, ResultMeasureValue, ResultMeasure.MeasureUnitCode)
```

## Overview

Trace metal concentrations are driven by geology, weathering, and, in some contexts, human activities. For the Nushagak watershed, baseline metal levels are important to characterize in light of mining interest in the region.

```{r metals_freq}
metal_freq <- metals_data %>%
  group_by(CharacteristicName) %>%
  summarize(
    Observations = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Observations)) %>%
  rename("Metal/Element" = CharacteristicName, "Measurements" = Observations)

knitr::kable(metal_freq, caption = "Frequency of Metal/Element Measurements")
```

## Copper

Copper is toxic to aquatic organisms at elevated concentrations and is particularly relevant given mining development scenarios. EPA freshwater aquatic life criteria: 13 µg/L (criterion continuous concentration).

```{r copper_summary}
# Get complete copper data with sample fraction information
wq_copper_full <- readWQPdata(huc = nushagak_hucs, characteristicName = "Copper", sampleMedia = "Water")

copper_data <- wq_copper_full %>%
  filter(!is.na(ResultMeasureValue)) %>%
  mutate(
    Value = as.numeric(ResultMeasureValue),
    Form = if_else(is.na(ResultSampleFractionText), "Unspecified", ResultSampleFractionText)
  ) %>%
  select(Form, Value) %>%
  filter(!is.na(Value))

if (nrow(copper_data) > 0) {
  copper_summary <- copper_data %>%
    group_by(Form) %>%
    summarize(
      N = n(),
      Mean = mean(Value, na.rm = TRUE),
      Median = median(Value, na.rm = TRUE),
      Min = min(Value, na.rm = TRUE),
      Max = max(Value, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    arrange(desc(N)) %>%
    rename("Sample Form" = Form, "Count" = N)
  
  knitr::kable(copper_summary, digits = 2, caption = "Copper Concentrations by Sample Form (µg/L)")
} else {
  cat("No copper data available in WQP.\n")
}
```

## Other Metals

Zinc, lead, cadmium, and chromium have also been measured at some locations. Distribution and frequency vary significantly.

```{r other_metals_table}
other_metals <- metals_data %>%
  filter(CharacteristicName %in% c("Zinc", "Lead", "Cadmium", "Chromium")) %>%
  mutate(Value = as.numeric(ResultMeasureValue)) %>%
  group_by(CharacteristicName) %>%
  summarize(
    N = n(),
    Mean = mean(Value, na.rm = TRUE),
    Min = min(Value, na.rm = TRUE),
    Max = max(Value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(N)) %>%
  rename("Element" = CharacteristicName, "Count" = N)

if (nrow(other_metals) > 0) {
  knitr::kable(other_metals, digits = 2, caption = "Selected Metals: Summary Statistics")
} else {
  cat("Limited data available for other metals in WQP.\n")
}
```

## Data Quality Notes

WQP returns metal concentration measurements as character strings, which must be explicitly converted to numeric type before statistical calculations. This analysis performs that conversion and filters missing values; however, because data submission practices vary, some observations may be incomplete or use unconventional units or detection limit notations. For larger monitoring efforts, the EPA's **TADA (Tools for Automated Data Analysis)** framework systematically handles these issues: validating metadata against EPA WQX domain standards, converting between units, and managing censored (below-detection-limit) observations. See the Data Management chapter for details.

## Data Gaps

Public metal concentration data in the Nushagak watershed are sparse; though non-public sources not covered yet in this analysis likely exist. Most observations come from targeted research projects rather than systematic monitoring. This represents a significant monitoring opportunity:

- **Dissolved vs. Total Metal**: Not all analyses distinguish between dissolved and total concentrations
- **Spatial Coverage**: Limited sampling outside established USGS gaging stations
- **Seasonal Variation**: Few multi-season samples; high-flow dynamics unknown
- **Speciation and Acid-Generating Potential**: No public baseline assessments in most tributaries

---

These gaps underscore the value of establishing baseline metal monitoring across the four Nushagak HUC8 sub-basins.

The next chapter examines nutrients and biological parameters relevant to salmon productivity.
