---
title: "Nushagak River Watershed: Water Quality Data Summary & Baseline Assessment"
author: "Prepared via Gemini for RStudio & Posit Assistant Migration"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    theme: cosmo
execute:
  warning: false
  message: false
---

## Project Context & Objectives

This document forms the core analytical component of a comprehensive water quality synthesis for the **Nushagak River Watershed** (HUC6: 190303) in southwestern Alaska.

### Core Project Goals:
1. **Data Aggregation**: Compile available water quality records across both major public repositories (EPA WQX/CDX, USGS NWIS) and non-repository literature (grey literature, regulatory baseline reports like the Pebble EIS EBD, and tribal monitoring).
2. **Spatial Assessment**: Map all historical and active sampling stations across the constituent HUC8 sub-basins:
   * **19030301** – Upper Nushagak River
   * **19030302** – Mulchatna River
   * **19030303** – Nuyakuk River
   * **19030304** – Wood River
   * **19030305** – Snake River / Lower Nushagak
3. **Workflow Migration**: Transition initial exploratory work from Gemini into a standalone **RStudio Project** to execute advanced data retrieval, parameter filtering (pH, dissolved oxygen, metals, nutrients), and statistical modeling with **Posit Assistant**.

---

## Setup & Dependencies

```{r setup}
# Required packages for API querying, spatial analysis, and dynamic reporting
required_pkgs <- c("dataRetrieval", "sf", "leaflet", "dplyr", "ggplot2", "DT", "tidyr")

# Install any missing dependencies automatically
new_pkgs <- required_pkgs[!(required_pkgs %in% installed.packages()[,"Package"])]
if(length(new_pkgs)) install.packages(new_pkgs)

# Load libraries
library(dataRetrieval) # USGS / WQP REST API interface
library(sf)            # Spatial data frame management
library(leaflet)       # Interactive spatial visualization
library(dplyr)         # Data manipulation
library(ggplot2)       # Data visualization
library(DT)            # Interactive HTML data tables
library(tidyr)         # Tidy data transformation
```

---

## 1. Public Repository Inventory Query

Using the `dataRetrieval` package, we query the Water Quality Portal (WQP) to build a site inventory across all five target HUC8 sub-basins without downloading full measurement payloads immediately.

```{r get_wqp_inventory}
# Target HUC8 basins in the Nushagak drainage
nushagak_hucs <- c("19030301", "19030302", "19030303", "19030304", "19030305")

# Fetch location metadata for all stations in these HUCs
wq_inventory <- whatWQPdata(huc = nushagak_hucs)

# Convert inventory to sf spatial object for mapping
wq_sf <- wq_inventory %>%
  filter(!is.na(dec_lon_va), !is.na(dec_lat_va)) %>%
  st_as_sf(coords = c("dec_lon_va", "dec_lat_va"), crs = 4326)
```

---

## 2. Watershed Interactive Site Map

The interactive map below displays monitoring stations retrieved from WQP across the Nushagak watershed. Click any marker to view station details and total recorded sample counts.

```{r site_map}
leaflet(wq_sf) %>%
  addProviderTiles(providers$Esri.WorldTopoMap, group = "Topo") %>%
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
  addCircleMarkers(
    radius = 5,
    color = "#1f78b4",
    stroke = TRUE,
    weight = 1,
    fillOpacity = 0.7,
    popup = ~paste0(
      "<b>Organization:</b> ", OrganizationFormalName, "<br>",
      "<b>Site Name:</b> ", MonitoringLocationName, "<br>",
      "<b>Site ID:</b> ", MonitoringLocationIdentifier, "<br>",
      "<b>Site Type:</b> ", ResolvedMonitoringLocationTypeName, "<br>",
      "<b>Result Count:</b> ", resultCount
    )
  ) %>%
  addLayersControl(
    baseGroups = c("Topo", "Satellite"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

---

## 3. Monitoring Effort by Organization

The table below summarizes total sampling locations and observation counts across providing organizations (e.g., USGS, EPA, Alaska DEC, Tribal entities).

```{r summarize_data}
org_summary <- wq_inventory %>%
  group_by(OrganizationFormalName) %>%
  summarize(
    Total_Sites = n_distinct(MonitoringLocationIdentifier),
    Total_Results_Available = sum(resultCount, na.rm = TRUE)
  ) %>%
  arrange(desc(Total_Results_Available))

datatable(org_summary, rownames = FALSE, 
          options = list(pageLength = 10, autoWidth = TRUE))
```

---

## 4. Next Steps: Parameter Downloading via Posit Assistant

To fetch actual parameter measurements (e.g., pH, metals, temperature, dissolved oxygen) for focused sub-basins like the Mulchatna River (HUC 19030302), run the script below in RStudio or prompt Posit Assistant.

```{r fetch_raw_data, eval=FALSE}
# Retrieve full surface water dataset for HUC 19030302
mulchatna_wq_data <- readWQPdata(
  huc = "19030302",
  sampleMedia = "Water"
)

# Extract key parameters of interest
key_params <- mulchatna_wq_data %>% 
  filter(CharacteristicName %in% c("pH", "Dissolved oxygen (DO)", "Temperature, water", "Copper", "Turbidity"))

# Parameter frequency table
table(key_params$CharacteristicName)
```
